---
title: "Exercise Analysis"
author: "Joseph Oberlander"
date: "February 10, 2017"
output: html_document
---
#### Background
(from coursera assignment)
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).
Data

The training data for this project are available here: [https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv]

The test data are available here: [https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv]

The data for this project come from this source: [http://groupware.les.inf.puc-rio.br/har]. 

If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.
What you should submit

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-). You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details.

#### Preliminary Work
An overall pseudo-random number generator seed was set at 697, so to reproduce these results, the same seed should be used. 

These packages are required: caret and randomForest.
### The model
Since the outcome variable is classe, it is a factor variable with 5 levels. For this data set, "participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions:

  # exactly according to the specification (Class A)
  # throwing the elbows to the front (Class B)
  # lifting the dumbbell only halfway (Class C)
  # lowering the dumbbell only halfway (Class D)
  # throwing the hips to the front (Class E)

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes." 

Predictions attempt to maximize accuracy and minimize the out-of-sample error, and available variables after cleaning will be used for predictions. Two models will be tested using decision tree and random forest algorithms, and the model with the highest accuracy will be chosen as our final model.

### Cross-validation

Cross-validation will be performed via subsampling of a training data, set randomly without replacement into 2 subsamples: 
  # subTraining data (75% of the original Training data set)
  # subTesting data (the remaining 25%). 
  
These models will be fitted on the subTraining data set, and tested on the subTesting data. Following selection of the most accurate model, it will be tested on the original Testing data set.

  # The expected out-of-sample error will relate to the quantity of data: 1-accuracy in the cross-validation data. 
  # Accuracy is defined as the proportion of correct classified observation over the total sample in the subTesting data set. 
  # Expected accuracy is the expected accuracy in the out-of-sample data set (i.e. original testing data set). 

Thus, the expected value of the out-of-sample error will correspond to the expected number of missclassified observations/total observations in the Test data set, which is the quantity: 1-accuracy found from the cross-validation data set.

The outcome variable "classe" is an unordered factor variable, so the error type can be chosen as 1-accuracy. With a large sample size (N= 19622) in the Training data set,thus cross-validation is allowed by dividing the Training sample into subTraining and subTesting.

Features with all missing values will be discarded as well as features that are irrelevant. All other features will be kept as relevant variables. Decision tree and random forest algorithms are known for their ability of detecting the features that are important for classification.

### Packages, Libraries and Seed
Install packages and load libraries, and set the seed for reproducability:
```{r, echo = TRUE}
#install.packages("caret")
library(caret)
#install.packages("randomForest")
library(randomForest)
#install.packages("rpart")
library(rpart) 
#install.packages("rpart.plot")
library(rpart.plot)
library(RColorBrewer)
#install.packages("rattle")
library(rattle)
set.seed(697)
```
### Fetching and cleaning data
The training data set is located at:
```{r, echo = TRUE}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
```
The testing data set is located at:
```{r, echo = TRUE}
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```
Fetch the data
```{r, echo = TRUE}
train <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
test <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```
Partioning the training set into two parts (60% for Training, 40% for Testing):
```{r, echo = TRUE}
inTrain <- createDataPartition(y=train$classe, p=0.6, list=FALSE)
Training <- train[inTrain, ]; Testing <- train[-inTrain, ]
dim(Training); dim(Testing)
```
### Cleaning the data
The data were cleaned with these transformations.
NearZeroVariance Variables were viewed
```{r,echo=TRUE}
myDataNZV <- nearZeroVar(Training, saveMetrics=TRUE)
myNZVvars <- names(Training) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt",
"kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
"max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
"var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
"stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
"kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
"max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
"kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
"skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
"amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
"skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
"max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
"amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
"avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
"stddev_yaw_forearm", "var_yaw_forearm")
myTraining <- Training[!myNZVvars]
```
First column of Dataset (ID) Removed for simplicity
```{r,echo=TRUE}
myTraining <- myTraining[c(-1)]
```
Variables more than a 60% "NA" removed:
```{r,echo=TRUE}
trainingV3 <- myTraining
for(i in 1:length(myTraining)) { 
        if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .6 ) {
        for(j in 1:length(trainingV3)) {
            if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) ==1)  {
                trainingV3 <- trainingV3[ , -j]
            }   
        } 
    }
}
myTraining <- trainingV3
rm(trainingV3)
```
The same 3 transformations for myTesting and testing data sets:
```{r,echo=TRUE}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58])
myTesting <- Testing[clean1]
testing <- Testing[clean2]
```

In order to ensure properly functioning Decision Trees and RandomForest Algorithms with the Test data set, the data is coerced into the same type.
```{r,echo=TRUE}
for (i in 1:length(testing) ) {
        for(j in 1:length(myTraining)) {
        if( length( grep(names(myTraining[i]), names(testing)[j]) ) ==1)  {
            class(testing[j]) <- class(myTraining[i])
        }      
    }      
}

testing <- rbind(myTraining[2, -58] , testing)
testing <- testing[-1,]
```
Using ML algorithms for prediction: Decision Tree
```{r,echo=TRUE}
modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
```
To view the decision tree with fancy :
```{r,echo=TRUE}
fancyRpartPlot(modFitA1)
```
Predicting:
```{r,echo=TRUE}
predictionsA1 <- predict(modFitA1, myTesting, type = "class")
```
Using ML algorithms for prediction: Random Forests
```{r,echo=TRUE}
modFitB1 <- randomForest(classe ~. , data=myTraining)
```
Predicting in-sample error:
```{r,echo=TRUE}
predictionsB1 <- predict(modFitB1, myTesting, type = "class")
```

Based on these models, the Random Forests yielded the best Results. Using the provided Test, out-of-sample error is set.
```{r,echo=TRUE}
predictionsB2 <- predict(modFitB1, testing, type = "class")
```
Function to generate files with predictions to submit for assignment
```{r,echo=TRUE}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(predictionsB2)
```
